In mathematics you don't understand things, you just get used to them."
~John von Neumann

"Study as if you were going to live forever; live as if you were going to die tomorrow."
~Maria Mitchell

 "Science is built up of facts, as a house is built of stones; but an accumulation of facts is no more a science than a heap of stones is a house."
~Henri Poincarķ

"My mother made me a scientist without ever intending to. Every other Jewish mother in Brooklyn would ask her child after school, 'So? Did you learn anything today?' But not my mother. 'Izzy,' she would say, 'did you ask a good question today?' That difference - asking good questions - made me become a scientist."
~Isidor Isaac Rabi

Reserve your right to think, for even to think wrongly is better than not to think at all."
~Hypatia

Science makes people reach selflessly for truth and objectivity; it teaches people to accept reality, with wonder and admiration, not to mention the deep awe and joy that the natural order of things brings to the true scientist."
~Lise Meitner

Not only is the Universe stranger than we think, it is stranger than we can think."
~Werner Heisenberg

Amicus Plato amicus Aristoteles magis amica verita. Plato is my friend, Aristotle is my friend, but my greatest friend is truth."
~Sir Isaac Newton

Why do children dread mathematics? Because of the wrong approach. Because it is looked at as a subject. 
Shakuntala Devi

As far as the laws of mathematics refer to reality, they are not certain, and as far as they are certain, they do not refer to reality. 
Albert Einstein

The essence of mathematics is not to make simple things complicated, but to make complicated things simple.
-S. Gudder

Go down deep enough into anything and you will find mathematics.
-Dean Schlicter

Trigonometry is a sine of the times.
-author unknow

There are 10 types of people in this world: those who understand binary and those who don't.
-author unknown

A mathematician is a blind man in a dark room looking for a black cat which isn't there.
-Charles R. Darwin

Science is simply common sense at its best.
-Thomas Huxley

Facts are not science - as the dictionary is not literature.
-Martin H. Fischer

Even if the open windows of science at first make us shiver after the cozy indoor warmth of traditional humanizing myths, in the end the fresh air brings vigor, and the great spaces have a splendor of their own.
-Bertrand Russell

Physics is mathematical not because we know so much about the physical world, but because we know so little; it is only its mathematical properties that we can discover.ö 
? Bertrand Russell

ōThis does not mean that science is just the art of making measurable predictions. Some philosophers of science overly circumscribe science by limiting it to its numerical predictions. They miss the point, because they confuse the instruments with the objectives. Verifiable quantitative predictions are instruments to validate hypotheses. The objective of scientific research is not just to arrive at predictions: it is to understand how the world functions; to construct and develop an image of the world, a conceptual structure to enable us to think about. Before being technical, science is visionary.ö 
? Carlo Rovelli, La realtÓ non Ķ come ci appare: La struttura elementare delle cose

ōScience does not stand as a moral guardian of humanity, rather it attempts to stand as the least subjective, least biased and least sociologically conditioned friend to humanity, when humanity seeks answers of real significance.ö 
? Abhijit Naskar, Morality Absolute

When Kepler found his long-cherished belief did not agree with the most precise observation, he accepted the uncomfortable fact. He preferred the hard truth to his dearest illusions; that is the heart of science.ö 
-Carl Sagan

People like us, who believe in physics, know that the distinction between past, present, and future is only a stubbornly persistent illusion.ö
-Albert Einstein

If we knew what it was we were doing, it would not be called research, would it?ö 
? Albert Einstein

Somewhere, something incredible is waiting to be known.ö 
? Carl Sagan

ōIn the beginning there was nothing, which exploded.ö 
? Terry Pratchett, Lords and Ladies

ōThat which can be asserted without evidence, can be dismissed without evidence.ö 
? Christopher Hitchens

Your assumptions are your windows on the world. Scrub them off every once in a while, or the light won't come in.ö 
? Isaac Asimov

IÆm so tired... I was up all night trying to round off infinity.ö 
? Steven Wright

A false conclusion once arrived at and widely accepted is not easily dislodged and the less it is understood the more tenaciously it is held.
Georg Cantor

